\documentclass[a4paper,11pt,oneside]{article}

\makeatletter
\def\input@path{{../}{}}
\makeatother
\usepackage{style/project}

% Exercise Sheet 03

\begin{document}

\begin{center}
    {\LARGE\bfseries Groups and Lie Algebras: Exercise Sheet 3\par}
    \vspace{0.5em}
    {\color{gray!70}\rule{0.72\textwidth}{0.5pt}\par}
\end{center}

\setcounter{section}{3}

    \ec{Some properties of characters}{
        Let \(V\) and \(W\) be a finite dimensional representation of a finite
        group \(G\) with characters \(\chi_V\) and \(\chi_W\). Compute the
        character of the representations
        \[
            V \oplus W,\quad
            V \otimes W,\quad
            V^*,\quad
            \operatorname{Hom}(V,W),\quad
            \Lambda^2 V,\quad
            \operatorname{Sym}^2 V.
        \]
    }{

    }

    \ecs{Quaternions}{
        The quaternion group consists of \(8\) elements
        \[
            \pm 1,\quad \pm i,\quad \pm j,\quad \pm k.
        \]
        The defining relations are
        \[
            i^2 = j^2 = k^2 = -1,\qquad
            i = jk = -kj,\qquad
            j = ki = -ik,\qquad
            k = ij = -ji.
        \]
    }{
        \ecpart{First find \(1\) four-dimensional representations of \(Q_8\).}{
            We start by writing the elements of \(Q_8\) as the vectors
            \[
                1 = \begin{pmatrix}1\\0\\0\\0\end{pmatrix},\quad
                i = \begin{pmatrix}0\\1\\0\\0\end{pmatrix},\quad
                j = \begin{pmatrix}0\\0\\1\\0\end{pmatrix},\quad
                k = \begin{pmatrix}0\\0\\0\\1\end{pmatrix}
            \]
            We can then define a representation by left multiplication. To find the explicit form we use the above relations. 1 acts as the identity, so we start by considering the action of \(i\):
            \[\rho(i) \cdot 1 = 1, \quad \rho(i) \cdot i = 1, \quad \rho(i) \cdot j = k, \quad \rho(i) \cdot k = -j \]
            so we have
            \[\rho(i) = \begin{pmatrix}
                0 & -1 & 0 & 0 \\
                1 & 0 & 0 & 0 \\
                0 & 0 & 0 & -1 \\
                0 & 0 & 1 & 0
            \end{pmatrix}\]
            We can do the same for \(j\) and \(k\):
            \[\rho(j) = \begin{pmatrix}
                0 & 0 & -1 & 0 \\
                0 & 0 & 0 & 1 \\
                1 & 0 & 0 & 0 \\
                0 & -1 & 0 & 0
            \end{pmatrix}, \quad
            \rho(k) = \begin{pmatrix}
                0 & 0 & 0 & -1 \\
                0 & 0 & -1 & 0 \\
                0 & 1 & 0 & 0 \\
                1 & 0 & 0 & 0
            \end{pmatrix}
            \]
            To check if this representation is irreducible, we compute the inner product of the character with itself:
            \[
                \langle \chi, \chi \rangle = \frac{1}{|Q_8|} \sum_{g \in Q_8} |\chi(g)|^2 = \frac18 (2\cdot4^2 + 6\cdot0^2) \neq 1
            \]
            This means this representation is not irreducible.
        }

        \ecpart{Find \(4\) one-dimensinal representations of \(Q_8\). Are they irreducible?}{
            To find the one-dimensional representations, we find a homomorphism from \(Q_8\) to \(\mathbb{C}^\times\). 
            We start by considering \(\rho(1) = 1 \) and \(\rho(-1)^2=\rho(1)=1\). This means \(\rho(-1) \in \{\pm1\}\). Similarly, we have \(\rho(i)^2 = \rho(j)^2 = \rho(k)^2 = \rho(-1)\). Thus if \(\rho(-1) = 1\) then we have \(\rho(i) = \rho(j) = \rho(k) \in \{ \pm1 \}\) and if \(\rho(-1) = -1\) then we have \(\rho(i), \rho(j), \rho(k) \in \{\pm i\}\).
            From the relations above we can exclude \(\rho(-1) = -1 \) since \(\rho(i) = \rho(j)\rho(k) = \pm i^2 \notin \{\pm i\}\). Thus we can write down all one-dimensional representations by choosing \(\rho(i), \rho(j), \rho(k) \in \{\pm1\}\). But since \(\rho(k) = \rho(i)\rho(j)\), only four choices are possible. We can write them explicitly in the following table:
            \[
                \begin{array}{c|ccccc}
                    & 1 & -1 & i & j & k\\
                    \hline
                    \rho_1 & 1 & 1 & 1 & 1 & 1 \\
                    \rho_2 & 1 & 1 & -1 & -1 & 1 \\
                    \rho_3 & 1 & 1 & -1 & 1 & -1 \\
                    \rho_4 & 1 & 1 & 1 & -1 & -1
                \end{array}
            \]
            They are all irreducible since they are one-dimensional.
        }

        \ecpart{There is also a \(2\)-dimensional irreducible representation.
        To find it, think of quantum mechanics.}{
            To find the 2 dimensional representation, we use the pauli matrices. The paulimatrices are unitary and self-adjoint, therefore we multiply them by \(i\) to get a representation of \(Q_8\):
            \[\rho(i) = i\sigma_1 =
            \begin{pmatrix}
                i & 0 \\
                0 & -i
            \end{pmatrix}, \quad
            \rho(j) = i\sigma_2 = 
            \begin{pmatrix}
                0 & 1 \\
                -1 & 0
            \end{pmatrix}, \quad
            \rho(k) = i\sigma_3 = 
            \begin{pmatrix}
                0 & i \\
                i & 0
            \end{pmatrix}
            \]
            Again we check if this representation is irreducible by computing the inner product of the character with itself:
            \[\langle \chi, \chi \rangle = \frac{1}{|Q_8|} \sum_{g \in Q_8} |\chi(g)|^2 = \frac18 (2\cdot2^2 + 6\cdot0^2) = 1\]
            This means this representation is irreducible.
        }

        \ecpart{Are these all irreducible representations? Explain.}{
            To check if wa have all irreducible representations, we use the following two facts:
            \begin{itemize}
                \item \(\mid \operatorname{Conj(G)}\mid = \sum_{\rho \,\text{irrep.}} 1\)
                \item \(|G| = \sum_{\rho \,\text{irrep.}} (\dim \rho)^2\)
            \end{itemize}
            First we compute the conjugacy classes of \(Q_8\):
            \[
                \operatorname{Conj}(1) = \{g\in Q_8 \mid g=q1q^{-1}\} = \{1\} \\
            \]
            \[
                \operatorname{Conj}(-1) = \{g\in Q_8 \mid g=q(-1)q^{-1}\} = \{-1\} \\
            \]
            Are trivialy given. From the defining relations we can also compute the conjugacy classes of \(i\), \(j\) and \(k\):
            \[
                i = jk \quad  j = ki \, \Longrightarrow \, -j = -ki = -kjk = kjk^{-1} \\
            \]
            Similarly we can compute all conjugations of \(i\), \(j\) and \(k\) and we find that
            \[\operatorname{Conj}(i) = \{i, -i\}, \quad \operatorname{Conj}(j) = \{j, -j\}, \quad \operatorname{Conj}(k) = \{k, -k\}\]
            So we have 5 conjugacy classes, which means we need 5 irreducible representations. Therefore the 4 one-dimensional representations and the 2-dimensional representation we found are all irreducible representations of \(Q_8\).

            We can also derive the fact that the irreducible representations of \(Q_8\) are four 1-dimensional and one 2-dimensional representation using the second fact above. We have
            \[|Q_8| = 8 = 2 \cdot 2^2 = 2^2 + 4 \cdot 1^2 = 8 \cdot 1^2\]
            Since we need five irreducible representations, we can only have four 1-dimensional and one 2-dimensional representation.
        }

        \ecpart{Compute the character table.}{
            We can now compute the character table of \(Q_8\) using the representations we found and the conjugacy classes:
            \[
                \begin{array}{c|ccccc}
                    & \{1\} & \{-1\} & \{\pm i\} & \{\pm j\} & \{\pm k\}\\
                    \hline
                    \rho_1 & 1 & 1 & 1 & 1 & 1 \\
                    \rho_2 & 1 & 1 & -1 & -1 & 1 \\
                    \rho_3 & 1 & 1 & -1 & 1 & -1 \\
                    \rho_4 & 1 & 1 & 1 & -1 & -1 \\
                    \rho_5 & 2 & -2 & 0 & 0 & 0
                \end{array}
            \]
        }
    }

    \ecs{The symmetric group \(S_4\) and its representations}{

    }{
        \ecpart{Determine the conjugacy classes of \(S_4\) and associate Young
        diagrams and conjugacy classes.}{

        }

        \ecpart{There are two \(1\)-dimensional representations of \(S_4\).
        What are they?}{

        }

        \ecpart{Consider the \(4\)-dimensional standard representation of
        \(S_4\) on \(\mathbb{R}^4\). Decompose it into a \(3\)-dimensional
        irreducible representation and a \(1\)-dimensional representation.}{

        }

        \ecpart{Take the tensor product of the \(3\)-dimensional representation
        of the previous part and the alternating representation. It is again
        irreducible. How many other irreducible representations are there? What
        are their dimensions?}{

        }

        \ecpart{Write down the character table.}{

        }
    }

    \ec{Particle in a box}{
        The Hilbert space of a particle in a box of length \(l\) is
        \(L^2([-l,l]^2)\) with the boundary conditions
        \[
            \psi|_{\partial[-l,l]^2} = 0
        \]
        and Hamiltonian
        \[
            \mathbb{H}
            =
            -\frac{\hbar^2}{2m}\left(\partial_x^2+\partial_y^2\right).
        \]
        Show that this system has \(D_4\) symmetry. Solve the Schr\"odinger
        equation and find an energy level whose eigenfunctions transform in the
        \(2\)-dimensional representation of \(D_4\). What is the lowest energy
        level that does not transform in an irreducible representation of
        \(D_4\).
    }{

    }

\end{document}
